Entropy-SGD: Biasing Gradient Descent Into Wide Valleys

### 2017-01-11 doc/chaudhari.choromanska.ea.iclr17.tex
minor

Based on this understanding of how the local geometry looks at the end of optimization, can we modify SGD to actively seek such regions? Motivated by the work of~\citet{baldassi2015subdominant} on shallow networks, instead of minimizing the original loss $f(x)$, we propose to maximize
$$
F(x, \g) = \log \int_{x' \in \reals^n}\ \exp \rbrac{-f(x') - \f{\g}{2}\ \norm{x-x'}^2_2}\ dx'.
$$
The above is a log-partition function that measures both the depth of a valley at a location  $x \in \reals^n$, and its flatness through the entropy of $f(x')$; we call it ``local entropy'' in analogy to the free entropy used in statistical physics. The $\entropysgd$ algorithm presented in this paper employs stochastic gradient Langevin dynamics (SGLD) to approximate the gradient of local entropy. Our algorithm resembles two nested loops of SGD: the inner loop consists of SGLD iterations while the outer loop updates the parameters. We show that the above modified loss function results in a smoother energy landscape defined by the hyper-parameter $\g$ which we can think of as a ``scope'' that seeks out valleys of specific widths. Actively biasing the optimization towards wide valleys in the energy landscape results in better generalization error. We present experimental results on fully-connected and convolutional neural networks (CNNs) on the MNIST and CIFAR-10~\citep{krizhevsky2009learning} datasets and recurrent neural networks (RNNs) on the Penn Tree Bank dataset (PTB)~\citep{marcus1993building} and character-level text prediction. Our experiments show that $\entropysgd$ scales to deep networks used in practice, obtains comparable generalization error as competitive baselines and also trains much more quickly than SGD (we get a $2\trm{x}$ speed-up for RNNs).

@@

Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the training loss and the width of local minima and show using the Gibbs formalism~\citep{haussler1997mutual} that this leads to improved generalization. As the authors discuss, the effect of hyper-parameters for the constraints they use is intricately tied together and they are difficult to choose even for small problems. Our local entropy based objective naturally balances the energetic term (training loss) and the entropic term (width of the valley) and also results in a more scalable algorithm. The scope $\g$ affords more control on the optimization process than hard constraints and as we discuss in Sec.~\ref{ss:scoping}, this provides significant computational advantages. Let us note that the Gibbs formalism introduced in~\citet{hochreiter1997flat} seems to be a very promising direction to understanding generalization in deep networks, it bears supercial similarities to the uniform stability analysis~\citep{haussler1997mutual} we employ. Recent work by~\citet{keskar2016large} is relevant in the context of the main theme of our paper. The authors estimate the loss in a neighborhood of the weights to argue that small batch size in SGD (i.e., larger gradient noise) generalizes better than large mini-batches and also results in significantly flatter minima.

\todo{In the sense that local entropy results in a smoother energy landscape (cf.\@ Sec.~\ref{ss:theoretical_properties}), it resembles continuation methods~\citep{allgower2012numerical} that convolve the loss function to solve sequentially refined optimization problems. These techniques however require a special class of loss functions where successively smaller convolutions result in minimizers that lie close to each other or closed-form expressions of smoothened loss functions~\citep{DBLP:conf/icml/HazanLS16,mobahi2015link}; these may not be possible for deep networks with isolated minima and highly-nonlinear loss functions. Our modified objective instead weighs each location by the local entropy in its neighborhood and does not require stringent assumptions on the objective, it is also applicable to full-scale deep networks.} %\pc{Let us also note that motivated from continuation,~\citet{gulcehre2016mollifying} insert annealed additive noise in the activation function to approximately convolve the original objective function.}{This paper is not related to our work, it is related to continuation. I don't see any reason to add this sentence}

>>>>

Based on this understanding of how the local geometry looks at the end of optimization, can we modify SGD to actively seek such regions? Motivated by the work of~\citet{baldassi2015subdominant} on shallow networks, instead of minimizing the original loss $f(x)$, we propose to maximize
$$
F(x, \g) = \log \int_{x' \in \reals^n}\ \exp \rbrac{-f(x') - \f{\g}{2}\ \norm{x-x'}^2_2}\ dx'.
$$
The above is a log-partition function that measures both the depth of a valley at a location  $x \in \reals^n$, and its flatness through the entropy of $f(x')$; we call it ``local entropy'' in analogy to the free entropy used in statistical physics. The $\entropysgd$ algorithm presented in this paper employs stochastic gradient Langevin dynamics (SGLD) to approximate the gradient of local entropy. Our algorithm resembles two nested loops of SGD: the inner loop consists of SGLD iterations while the outer loop updates the parameters. We show that the above modified loss function results in a smoother energy landscape defined by the hyper-parameter $\g$ which we can think of as a ``scope'' that seeks out valleys of specific widths. Actively biasing the optimization towards wide valleys in the energy landscape results in better generalization error. We present experimental results on fully-connected and convolutional neural networks (CNNs) on the MNIST and CIFAR-10~\citep{krizhevsky2009learning} datasets and recurrent neural networks (RNNs) on the Penn Tree Bank dataset (PTB)~\citep{marcus1993building} and character-level text prediction. Our experiments show that $\entropysgd$ scales to deep networks used in practice, obtains comparable generalization error as competitive baselines and also trains much more quickly than SGD (we get a $2\trm{x}$ speed-up on RNNs).

@@

Motivated by the same final goal, viz., flat local minima, the authors in~\citet{hochreiter1997flat} introduce hard constraints on the training loss and the width of local minima and show using the Gibbs formalism~\citep{haussler1997mutual} that this leads to improved generalization. As the authors discuss, the effect of hyper-parameters for the constraints they use is intricately tied together and they are difficult to choose even for small problems. Our local entropy based objective naturally balances the energetic term (training loss) and the entropic term (width of the valley) and also results in a more scalable algorithm. The scope $\g$ affords more control on the optimization process than hard constraints and as we discuss in Sec.~\ref{ss:scoping}, this provides significant computational advantages. Let us note that the Gibbs formalism used in the analysis of~\citet{hochreiter1997flat} seems to be a very promising direction to understanding generalization in deep networks.
%, it bears supercial similarities to the uniform stability analysis~\citep{haussler1997mutual} we employ.
%Recent work by~\citet{keskar2016large} is relevant in the context of the main theme of our paper. The authors estimate the loss in a neighborhood of the weights to argue that small batch size in SGD (i.e., larger gradient noise) generalizes better than large mini-batches and also results in significantly flatter minima.

In the sense that local entropy results in a smoother energy landscape (cf.\@ Sec.~\ref{ss:theoretical_properties}), it is related to continuation methods~\citep{allgower2012numerical} that convolve the loss function to solve sequentially refined optimization problems~\citep{DBLP:conf/icml/HazanLS16,mobahi2015link} or methods such as~\citet{gulcehre2016mollifying} that \pc{perturb the weights with random noise to average the gradient}{cite some other paper instead}. Smoothing the

More principled techniques based on continuation

These techniques however require a special class of loss functions where successively smaller convolutions result in minimizers that lie close to each other or closed-form expressions of smoothened loss functions; these may not be possible for deep networks with isolated minima and highly-nonlinear loss functions.

Our modified objective instead weighs each location by the local entropy in its neighborhood and does not require stringent assumptions on the objective, it is also applicable to full-scale deep networks.
